---
title: "Introduction to StatComp"
author: "Hong Zhang"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to StatComp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Overview

__StatComp__ is a simple R package developed to compare the performance of R and R++ (implemented through the R package _Rcpp_) for the 'Statistical Computing' course. Two functions are considered, namely, _gibbs_ (generate random nubers using Gibbs sampler) and _vacc_ (predicting the response using three variables: age, gender, and ily). For each function, both R and Rcpp versions are produced. Namely _gibbsR_ and  _vaccR_ for R and 'gibbC' and _vaccC_ for C++.

The R package 'microbenchmark' can be used to benchmark the above R and C++ functions.

## Benchmarking _vaccR_ and _vaccC_

The source R code for _vaccR_ is as follows:
```{r,eval=FALSE}
function (age, female, ily) 
{
    p <- 0.25 + 0.3 * 1/(1 - exp(0.04 * age)) + 0.1 * ily
    p <- p * ifelse(female, 1.25, 0.75)
    p <- pmax(0, p)
    p <- pmin(1, p)
    p
}
```

The above code involves 'ifelse', 'pmax', and 'pmin', which are known to be slow. On the other hand, the following _Rcpp_ code is much faster.

```{r,eval=FALSE}
double vacc3a(double age, bool female, bool ily){
  double p = 0.25 + 0.3 * 1 / (1 - exp(0.04 * age)) + 0.1 * ily;
  p = p * (female ? 1.25 : 0.75);
  p = std::max(p, 0.0);
  p = std::min(p, 1.0);
  return p;
}
NumericVector vaccC(NumericVector age, LogicalVector female,
                    LogicalVector ily) {
  int n = age.size();
  NumericVector out(n);
  for(int i = 0; i < n; ++i) {
    out[i] = vacc3a(age[i], female[i], ily[i]);
  }
  return out;
}
```

In order to empirically benchmark _vaccR_ and _vaccC_, one generates 1,000 repicates of (age, female, ily), and save it in data{StatComp}. To load the file, one simply calls _data(data)_. The R code for benchmark _vaccR_ and _vaccC_ is as follows.

```{r,eval=TRUE}
library(StatComp)
library(microbenchmark)
data(data)
attach(data)
tm2 <- microbenchmark(
  vR = vaccR(age,female,ily),
  vC = vaccC(age,female,ily)
)
knitr::kable(summary(tm2)[,c(1,3,5,6)])
```

The above results show an evident computational speed gain of C++ against R.


## Benchmarking _gibbsR_ and _gibbsC_

The source R code for _vaccR_ is as follows:
```{r,eval=FALSE}
gibbsR <- function(N, thin) {
  mat <- matrix(nrow = N, ncol = 2)
  x <- y <- 0
  for (i in 1:N) {
    for (j in 1:thin) {
      x <- rgamma(1, 3, y * y + 4)
      y <- rnorm(1, 1 / (x + 1), 1 / sqrt(2 * (x + 1)))
    }
    mat[i, ] <- c(x, y)
  }
  mat
}
```

The above R code involves two loops, which could be very slow even for R-3.01 or any higher version. The corresponding C++ code is as follows.

```{r,eval=FALSE}
NumericMatrix gibbsC(int N, int thin) {
  NumericMatrix mat(N, 2);
  double x = 0, y = 0;
  for(int i = 0; i < N; i++) {
    for(int j = 0; j < thin; j++) {
      x = rgamma(1, 3, 1 / (y * y + 4))[0];
      y = rnorm(1, 1 / (x + 1), 1 / sqrt(2 * (x + 1)))[0];
    }
    mat(i, 0) = x;
    mat(i, 1) = y;
  }
  return(mat);
}
```

The R code for benchmarking _gibbsR_ and _gibbsC_ is as follows.

```{r,eval=TRUE}
tm2 <- microbenchmark(
  vR = gibbsR(1e4, 10),
  vC = gibbsC(1e4, 10)
)
knitr::kable(summary(tm2)[,c(1,3,5,6)])
```

The results again show an evident computational speed gain of C++ against R.

